---
title: "NFL Fast R Beginner Dic"
author: "Jeffrey Gordon"
format: gfm
editor: visual
---

```{r}
#| echo: false
library(tidyverse)
library(ggrepel)
library(nflreadr)
library(nflplotR)
library(gt)
library(ggthemes)
options(nflreadr.verbose = FALSE)
options(dplyr.summarise.inform = FALSE)
```

I am going to take a look at the progression of expected points added over the course of the season for the New York Jets offense. I want to break the season into two known chunks: quarterback at the time, and main running back.

```{r}
#| echo: false
jets_pbp <- load_pbp(2022) %>%
  filter(!is.na(posteam), 
         posteam == "NYJ", 
         season_type == "REG", 
         play_type == "pass" | play_type == "run"
         )
```

```{r}
#| echo: false
# view(jets)
jets_pbp %>%
  select(play_type, week) %>%
  # count(play_type, week) %>%
  group_by(play_type, week) %>%
  summarize(count_by_type = n()) %>%
  ggplot2::ggplot(aes(x=week,y = count_by_type, color=play_type)) + 
  geom_line() + geom_point() + theme_few() + scale_fill_few() +
  scale_x_continuous(breaks = 1:max(jets_pbp$week)) + 
  labs(x = "Week #", y = "# of Plays", title = "NYJ Offensive Plays by Type 2022")

```

```{r}
#| echo: false
jets_pbp %>%
  select(play_type, week) %>%
  # count(play_type, week) %>%
  group_by(play_type, week) %>%
  summarize(count_by_type = n()) %>%
  pivot_wider(names_from = week,
              values_from = count_by_type) %>%
  gt(groupname_col = "group") %>%
  tab_header(
    title = "Plays by Type 2022"
  )
  
```

As of the end of week 9 of the 2022 season, the New York Jets have run 561 offensive plays classified as a run or pass according to the nflfastR play-by-play data.

# EPA

What I want to explore next has to do with expected points added (EPA). For more information on how this is calculated you can visit:

-   [nflWAR: A Reproducible Method for Offensive Player Evaluation in Football (Extended Edition)](https://arxiv.org/pdf/1802.00998.pdf)

-   [Calibration for nflfastR models (including expected points)](https://www.opensourcefootball.com/posts/2020-09-28-nflfastr-ep-wp-and-cp-models/#ep-model-features)

I am going to explore the changing expected points over the course of the season.

```{r}
jets_pbp %>%
  filter(play_type == "pass") %>%
  select(play_id, play_type, week, epa) %>%
  ggplot2::ggplot(aes(x = factor(week), y = epa, fill = factor(play_type))) +
  geom_boxplot() + geom_jitter(width=0.1, alpha = 0.2) + theme_few()
```

```{r}
jets_pbp %>%
  filter(play_type == "run") %>%
  select(play_id, play_type, week, epa) %>%
  ggplot2::ggplot(aes(x = factor(week), y = epa, fill = factor(play_type))) +
  geom_boxplot() + geom_jitter(width=0.1, alpha = 0.2)
```

## Citations

Carl S, Baldwin B (2022). *nflfastR: Functions to Efficiently Access NFL Play by Play Data*. https://www.nflfastr.com/, https://github.com/nflverse/nflfastR.
